Statically typed Web programming with JavaScript

### 2012-11-23 papers/ecoop2013/repairing.tex
Update (but still not finished yet)

Actually, very few developers use the native API for creating DOM, another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:

@@

This implementation may be more readable, however the code is wrong: if we call \code{hello("<me>")} the brackets won’t be escaped and will produce a \code{ME} tag, which is not the expected behavior. So, although this way reads better it’s not really less error prone.

@@

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes and developers have the full computational power of Scala to inject dynamic data, DOM fragments definitions are written using functions so they compose just as functions compose.

@@

\begin{lstlisting}[language=JavaScript,label=efficient-null,caption=Defensive programming to handle null references]
var loginWidget = document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton = loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.addEventListener("click", function (e) { ... });
  }
}
\end{lstlisting}

@@

We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{efficient-null}. However, the checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.

Our solution has both the safety and performance of the listing \ref{efficient-null} and the expressiveness of the listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the \code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we just check if a value is \code{null} or not. Here is a listing using our DSL:

@@

To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function (cf listing~\ref{state-encapsulated}). Thus the state would be hidden and developers could not forget the initialization.

@@

The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability}. Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result. However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one:

@@

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across functions, as shown in the following listing:

@@

The State monad programming model has the advantages of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable and everything is immutable, but it is less verbose than these listings.

Our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.

@@

The code looks synchronous however the \code{sleep} calls are asynchronous: they translate to the following JavaScript:
\begin{lstlisting}[language=JavaScript]
window.setTimeout(function () {
  console.log("Hello");
  window.setTimeout(function () {
    console.log(", World!");
  }, 1000);
}, 1000);
\end{lstlisting}

@@

Instead, we propose a monadic DSL that does not attempt to hide its asynchronous nature but provides a way to lift asynchronous values as first-class citizen and, thanks to the \code{for} notation, a convenient syntax to avoid the callback-hell.

@@

Because callbacks are explicitly passed to the \code{foreach}, \code{map} and \code{flatMap} methods, there is no risk of ambiguity such as the one of listing \ref{async-ambiguous}. The Scala \code{for} notation handles the callback-hell problem.

@@

% TODO mentionner le traitement de cas d’erreurs ?

@@

@@

Kossakowski \etal{} showed how to use Scala classes to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes. However their solution suffers from some limitations: it doesn’t support open extension, it looses information on the side-effects performed by a method call (so it prevents some optimizations to be applied), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.

We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting open extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.

@@

// Main program
def main(user: Rep[User], point: Rep[Point]) = {
  println(show(user))
  println(show(point))
}
\end{lstlisting}
The generated code looks like the following:
\begin{lstlisting}[language=JavaScript]
var main = function (user, point) {
  var x0 = showUser(user);
  console.log(x0);
  var x1 = showPoint(point);
  console.log(x1);
};
var showUser = function (user) {
  var x2 = user.name + "(";
  var x3 = x2 + user.age;
  var x4 = x3 + " years)";
  return x4
};
var showPoint = function (point) {
  var x5 = "(" + point.x;
  var x6 = x5 + ", ";
  var x7 = x6 + point.y;
  var x8 = x7 + ")";
  return x8
};
\end{lstlisting}
Because the dispatch is performed at compile-time, the JavaScript code calls the right function according to its parameters, even though there is no way, in JavaScript, to get back the type of these parameters.

@@

What did we do?

>>>>

Another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:

@@

This implementation may be more readable, however the code is wrong: if we call \code{hello("<me>")} the brackets won’t be escaped and will produce a \code{ME} tag, which is not the expected behavior. So, although this way reads better it’s not less error prone.

@@

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and DOM fragments definitions are written using functions so they compose just as functions compose.

@@

\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget = document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton = loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.addEventListener("click", function (e) { ... });
  }
}
\end{lstlisting}

@@

We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{null-defensive}. However, the checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.

Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the \code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we just check if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code equivalent to the listing \ref{null-defensive}:

@@

To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function. Thus the state is hidden and developers may not forget the initialization:

@@

The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result:

@@

However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one. Furthermore, manually passing the result of each previous call to the next one is error prone.

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across computations, as shown in the following listing:

@@

The State monad programming model has the advantages but not the shortcomings of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives) and everything is immutable.

Furthermore, our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.

@@

The code looks synchronous however the \code{sleep} calls are asynchronous: they translate to the following JavaScript:

\begin{lstlisting}[language=JavaScript,label=async-js,caption=Generated code for listing \ref{async-continuation}]
window.setTimeout(function () {
  console.log("Hello");
  window.setTimeout(function () {
    console.log(", World!");
  }, 1000);
}, 1000);
\end{lstlisting}

@@

In this paper we propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen. Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.

@@

Because callbacks are explicitly passed to the \code{foreach}, \code{map} and \code{flatMap} methods, there is no risk of ambiguity such as the one of listing \ref{async-ambiguous}.

@@

The above listing produces the following code (we renamed some identifiers for the sake of readability):

\begin{lstlisting}[language=JavaScript]
var x1 = new Promise();
AjaxGet("/foo", x1);
var x2 = new Promise();
AjaxGet("/bar", x2);
x1.foreach(function (foo) {
  x2.foreach(function (bar) {
    var x3 = foo + bar;
    console.log(x3);
  });
});
\end{lstlisting}

The \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.

\begin{lstlisting}
def f(aF: Rep[Future[String]]) =
  for (a <- aF) yield a.toUpperCase
def g(aF: Rep[Future[String]]) =
  for (a <- aF) yield a.toLowerCase

def fAndG(aF: Rep[Future[String]]) = for {
  u <- f(aF)
  l <- g(aF)
} yield u + l
\end{lstlisting}

(TODO Explain the value of the above listing...)

@@

The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the operand coercion between heterogeneous types:

\begin{lstlisting}
def main(i: Rep[Int], j: Rep[Int],
         x: Rep[Double], y: Rep[Double]) = {
  println(i + j)
  println(x + y)
  // i is promoted to Rep[Double], n has type Rep[Double]
  val n = i + y
  // j is promoted to Rep[Double], m has type Rep[Double]
  val m = x + j
  println(n + m)
}
\end{lstlisting}

@@

Kossakowski \etal{} showed how to use Scala traits to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes. However their solution suffers from some limitations: it doesn’t support retroactive extension, it does not track the side-effects potentially performed by a method call (so it prevents some optimizations to be applied by LMS), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.

We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.

@@

// Main program
def main(user: Rep[User], point: Rep[Point]) = {
  println(show(user))
  println(show(point))
}
\end{lstlisting}
The generated code looks like the following:
\begin{lstlisting}[language=JavaScript]
var main = function (user, point) {
  var x0 = showUser(user);
  console.log(x0);
  var x1 = showPoint(point);
  console.log(x1);
};
var showUser = function (user) {
  var x2 = user.name + "(";
  var x3 = x2 + user.age;
  var x4 = x3 + " years)";
  return x4
};
var showPoint = function (point) {
  var x5 = "(" + point.x;
  var x6 = x5 + ", ";
  var x7 = x6 + point.y;
  var x8 = x7 + ")";
  return x8
};
\end{lstlisting}

(TODO Conclusion)

@@

We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
